---
title: "US_Ass_Juwas"
author: "Hans-Peter Bakker (BKKHAN001)"
date: "8/10/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Principal Components Analysis

```{r readingData}
# read in the data
j_dat <- read.table("leukemia_array.txt")

# transform the set turning rows into columns and vv
j_dat_tr <- data.frame(t(j_dat))
```

#### Initial Exploration of the Dataset

Before responding the to specific requirements of the assignment, an initial exploration of the dataset was undertaken. This exploration will serve to better inform the results and the conclusions that will be developed below.

```{r explr_samples}
#consider some samples
j_dat_tr[1,1:10]
j_dat_tr[9,1:10]

# consider some stats of the samples
par(mfrow = c(2,2))
sample_means <- apply(j_dat_tr, 1, mean)
plot(sample_means, col = c(rep(3,8), rep(2,8)), pch = 19)
sample_sds <- apply(j_dat_tr, 1, sd)
plot(sample_sds, col = c(rep(3,8), rep(2,8)), pch = 19)
sample_medians <- apply(j_dat_tr,1, median)
plot(sample_medians, col = c(rep(3,8), rep(2,8)), pch = 19)
```

The samples do not appear to differ much in terms of their means nor their standard deviations. There does however appear to be a systematic difference in their medians, with the "good" samples suggesting lower spread of medians than the "poor" samples. 

```{r expl_vars}
# consider distribution of the samples
par(mfrow = c(2,2))
hist(j_dat$p1_good)
hist((scale(j_dat$p1_good)))
hist(log(j_dat$p1_good))
hist(scale(log(j_dat$p1_good)))
```

```{r}
# consider some distributions of variables (gene expressions)
var_means <- apply(j_dat_tr, 2, mean)
hist(var_means)
var_sds <- apply(j_dat_tr,2, sd)
hist(var_sds)
var_medians <- apply(j_dat_tr,2, median)
hist(var_medians)
```

```{r}
hist(j_dat_tr$`1405_i_at`)
hist(scale(j_dat_tr$`1405_i_at`))
hist(scale(log(j_dat_tr$`1405_i_at`)))
```

```{r}
hist(var_means)
hist(var_sds)
hist(var_medians)

range(var_sds)
boxplot(var_sds)
```

### PCA: Original Data
```{r}
# doing pca on the unscaled data
pca_unscaled <- prcomp(x = j_dat_tr, center = FALSE, scale. = FALSE, retx = TRUE)
screeplot(pca_unscaled, type = "lines",npcs = 20)
sum_unscaled <- summary(pca_unscaled)
plot(sum_unscaled$importance[3,])
```

### PCA on Standardised Data
```{r}
# doing pca on scaled data
pca_scaled <- prcomp(j_dat_tr, center = TRUE, scale. = TRUE, retx = TRUE)
screeplot(pca_scaled, type = "lines")
sum_scaled <- summary(pca_scaled)
plot(sum_scaled$importance[3,])
```

### PCA on log_transformed Original Data
```{r}
# log-transforming the original dataset
j_dat_log <- as.data.frame(log(t(j_dat)))

# doing pca on log transformed data
pca_log <- prcomp(j_dat_log, center = FALSE, scale. = FALSE, retx = TRUE)
screeplot(pca_log, type = "lines")
sum_log <- summary(pca_log)
plot(sum_log$importance[3,])
```

### PCA on log-transformed and standardised data
```{r}
# doing same on log-transformed but standardised data
pca_log_tr <- prcomp(j_dat_log, center = TRUE, scale. = TRUE, retx = TRUE)
screeplot(pca_log_tr, type = "lines")
sum_log_tr <- summary(pca_log_tr)
plot(sum_log_tr$importance[3,])
```

### PCA on a reduced dataset
```{r}
# consider variance across samples for each gene expression
vars <- apply(j_dat_tr, 2, FUN = var)
top_100 <- sort(vars, decreasing = TRUE)[1:100]
ind_top_100 <- which(vars %in% top_100)
j_100 <- j_dat_tr[,ind_top_100]
```

### PCA on scaled and log-transformed set
```{r}
pca_100_logscaled <- prcomp(j_100, center = TRUE, scale. = TRUE, retx = TRUE)
screeplot(pca_100_logscaled, type = "lines")
sum_100_logscaled <- summary(pca_100_logscaled)
plot(sum_100_logscaled$importance[3,])
```

some thoughts:
consider variance of these 100 variables as percentage of total variance:
```{r}
vars_total <- sum(vars)
vars_100 <- sum(vars[ind_top_100])
prop_vars_100 <- vars_100/vars_total
```
represents 49.24% of total variance.
And from screeplot above, would consider elbow at 5 principal components. What is the variance explained by these 5?

```{r}
sum((pca_100_logscaled$sdev[1:5])^2)

```

compare this with percentage of total variance explained by the first 100 principal components (using the same log-transformed and then scaled) pca from above

```{r}

```
# if get rid of highly correlated variables...use caret and ...@0.9
# takes time...do when can...



#First log transform the data and then scale (think about value of log-transform...makes for more difficult interpretation, but does it add value..)
nu_j <- as.data.frame(scale(log(j_dat_tr)))

hist(nu_j$`1405_i_at`, breaks = 10)
hist(nu_j$`200019_s_at`, breaks = 10)
hist(nu_j$`202148_s_at`, breaks = 10)

#considering correlation of genes
#j_cor <- cor(j_dat_tr)
#library(caret)
#cor_08 <- findCorrelation(j_cor, cutoff = 0.8) # 20 000+
#cor_95 <- findCorrelation(j_cor, cutoff = 0.95) # 3802
# appears to be many strong correlations


